---
title: "Calculating Non-Dementia Mortality"
author: "Sigal Maya"
date: "Updated 02-20-2025"
output: 
  html_notebook: 
    fig_width: 3
    fig_height: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
getwd()
```

The model utilizes the risk ratio for mortality to evaluate the increased risk for death among those with dementia. However, using the risk ratio in conjunction with all-cause mortality inflates overall mortality in the model. We thus estimated the non-dementia-related mortality rate (i.e, mortality among those without the risk factor) by subtracting dementia-related deaths from all-cause deaths at each age. All-cause and dementia-related death rates were obtained from the CDC Wonder database.

```{r load}
path_allcause <- "data/mortality/allcause_deaths_bysex_byage.csv"
path_dementia <- "data/mortality/dementia_deaths_bysex_byage.csv"


allcause_raw <- read.csv(path_allcause)  
dementia_raw <- read.csv(path_dementia)  
```

```{r clean}
summary(allcause_raw)

allcause <- allcause_raw %>%
  janitor::clean_names() %>%
  select(-notes, -sex, -single_year_ages, -crude_rate) %>%
  rename(sex = sex_code,
         age = single_year_ages_code) %>%
  mutate(age = as.numeric(age),
         sex = as.factor(sex),
         population = as.numeric(population),
         rate = deaths/population) %>%
  select(-deaths, -population) %>%
  subset(age >= 50) %>%
  pivot_wider(names_from = sex, values_from = rate)

summary(allcause)  
# all converted to numeric, no NAs in age, 16 in rate (those aged 85+), all rates are non-zero

summary(dementia_raw)

dementia <- dementia_raw %>%
  janitor::clean_names() %>%
  select(-notes, -sex, -single_year_ages, -crude_rate) %>%
  rename(sex = sex_code,
         age = single_year_ages_code) %>%
  mutate(age = as.numeric(age),
         sex = as.factor(sex),
         population = as.numeric(population),
         rate = deaths/population) %>%
  select(-deaths, -population) %>%
  subset(age >= 50) %>%
  pivot_wider(names_from = sex, values_from = rate)

summary(dementia) 
# all converted to numeric, no NAs in age, 16 in rate (those age 85+), all rates are non-zero
```

Rates for older ages (85+) were unavailable, so we extrapolated using available data. First, we visually inspected the data to examine the distribution by age, confirming that the rate grew exponentially with age:

```{r plot original, include = TRUE}

par(mfrow = c(2, 2))
plot(allcause$age, allcause$`F`, xlab = "Age", ylab = "Rate", main = "Crude All-Cause\nMortality Rate by Age, Females")
plot(dementia$age, dementia$`F`, xlab = "Age", ylab = "Rate", main = "Crude Dementia-Related\nMortality Rate by Age, Females") 

plot(allcause$age, allcause$`M`, xlab = "Age", ylab = "Rate", main = "Crude All-Cause\nMortality Rate by Age, Males")
plot(dementia$age, dementia$`M`, xlab = "Age", ylab = "Rate", main = "Crude Dementia-Related\nMortality Rate by Age, Males") 


```

We used a log-transformed linear model with age as the predictor variable to extrapolate all-cause mortality rates up to the age of 100 (the maximum age allowed in the microsimulation). Initial models using all data points resulted in predictions that were not smooth, thus we used rates for ages 76 and up for predictions. The `predict()` function in R was used to predict the missing estimates, which were then exponentiated to obtain rates. Data were visually inspected to confirm a smooth and uninterrupted exponential increase in all-cause mortality rate with age.

```{r predict all-cause mortality rates}
any(allcause$`F`==0, na.rm=TRUE) # check to see if any zeros (log(0) is problematic) -- there are none
any(allcause$`M`==0, na.rm=TRUE) # no zeros
any(dementia$`F`==0, na.rm=TRUE) # no zeros
any(dementia$`M`==0, na.rm=TRUE) # no zeros


# Extrapolate all-cause mortality 
summary(allcause_log_f <- lm(log(`F`) ~ age, data = na.omit(allcause[allcause$age > 75, c("age","F")])))   
# using all data points gave weird fits, so only using ages 76 and up to predict
# good fit (R^2 = 0.99), p(age) ~ 0

summary(allcause_log_m <- lm(log(`M`) ~ age, data = na.omit(allcause[allcause$age > 75, c("age","M")])))   

allcause_predict <- data.frame(age = allcause$age[is.na(allcause$`M`)])
allcause_predict$f_predicted <- exp(predict(allcause_log_f, newdata = allcause_predict))

allcause_predict$m_predicted <- exp(predict(allcause_log_m, newdata = allcause_predict))

allcause_clean <- allcause %>%
left_join(allcause_predict, by = "age") %>% 
mutate(rate_f = coalesce(`F`, f_predicted),
       rate_m = coalesce(`M`, m_predicted)) %>%
select(-f_predicted, -m_predicted, -`F`, -`M`)
summary(allcause_clean)  # no NAs in rates

saveRDS(allcause_clean, "allcause_mortality_bysex_clean.RDS")
```

```{r all-cause mortality rates, include = TRUE}
par(mfrow = c(1, 2))

plot(allcause_clean$age, allcause_clean$rate_m, xlab = "Age", ylab = "Rate", main = "Extrapolated All-Cause Mortality Rate by Age, Males") # smooth exponential curve
plot(allcause_clean$age, allcause_clean$rate_f, xlab = "Age", ylab = "Rate", main = "Extrapolated All-Cause Mortality Rate by Age, Females")
```

Next, we calculated non-dementia-related mortality rates by subtracting dementia-related mortality rates from all-cause mortality rates. We fit another log-transformed linear model to the non-dementia-related mortality rates to obtain missing values for ages 85 and up.

```{r predict non-dementia mortality rates}
non_dementia <- allcause_clean[ ,c("age","rate_f","rate_m")] %>%
  right_join(dementia[ ,c("age","F","M")], by="age") %>%
  rename(f_rate_allcause = rate_f,
         f_rate_dementia = `F`,
         m_rate_allcause = rate_m,
         m_rate_dementia = `M`) %>%
  mutate(f_rate_non_dementia = pmax(f_rate_allcause - f_rate_dementia),
         m_rate_non_dementia = pmax(m_rate_allcause - m_rate_dementia))

summary(non_dementia)

summary(non_dementia_logf <- lm(log(f_rate_non_dementia) ~ age, data = na.omit(non_dementia[non_dementia$age>=76,])))
# good fit (R^2 ~= 1), p(age) ~ 0

summary(non_dementia_logm <- lm(log(m_rate_non_dementia) ~ age, data = na.omit(non_dementia[non_dementia$age>=76,])))


non_dementia_predict <- data.frame(age = non_dementia$age[is.na(non_dementia$m_rate_non_dementia)])
non_dementia_predict$f_rate_predicted <- exp(predict(non_dementia_logf, newdata = non_dementia_predict))  
non_dementia_predict$m_rate_predicted <- exp(predict(non_dementia_logm, newdata = non_dementia_predict))  


non_dementia_clean <- non_dementia %>%
left_join(non_dementia_predict, by = "age") %>% 
mutate(f_rate_non_dementia = coalesce(f_rate_non_dementia, f_rate_predicted),
       m_rate_non_dementia = coalesce(m_rate_non_dementia, m_rate_predicted)) %>%
select(age, f_rate_allcause, m_rate_allcause, f_rate_non_dementia, m_rate_non_dementia) 
summary(non_dementia_clean)
```

```{r non-dementia mortality rates, include = TRUE}
par(mfrow = c(1, 2))
plot(non_dementia_clean$age, non_dementia_clean$m_rate_non_dementia, xlab = "Age", ylab = "Rate",
     main = "Non-Dementia-Related Mortality Rate by Age, Males")

plot(non_dementia_clean$age, non_dementia_clean$f_rate_non_dementia, xlab = "Age", ylab = "Rate",
     main = "Non-Dementia-Related Mortality Rate by Age, Females") 

# smooth exponential curves

```

Finally, we converted rates to probabilities using the equation $\text{probability} = 1 - e^{-\text{rate}}$.

```{r convert rates to probabilities}
non_dementia_clean <- non_dementia_clean %>%
  mutate(m_prob_allcause = 1-exp(-m_rate_allcause),
         f_prob_allcause = 1-exp(-f_rate_allcause),
         m_prob_non_dementia = 1-exp(-m_rate_non_dementia),
         f_prob_non_dementia = 1-exp(-f_rate_non_dementia))
```

```{r inspect data}
non_dementia_grouped <- non_dementia_clean %>%
  mutate(age_group = cut(age, breaks = seq(50, 100, by = 5), right = FALSE)) %>%
  group_by(age_group) %>%
  summarise(f_avg_rate_allcause = mean(f_rate_allcause),
            f_avg_rate_non_dementia = mean(f_rate_non_dementia),
            m_avg_rate_allcause = mean(m_rate_allcause),
            m_avg_rate_non_dementia = mean(m_rate_non_dementia)) %>%
  mutate(
    m_avg_prob_allcause = 1-exp(-m_avg_rate_allcause),
    m_avg_prob_non_dementia = 1-exp(-m_avg_rate_non_dementia),
    f_avg_prob_allcause = 1-exp(-f_avg_rate_allcause),
    f_avg_prob_non_dementia = 1-exp(-f_avg_rate_non_dementia),
  )
```

```{r save data}
saveRDS(non_dementia_clean, "non_dementia_mortality_prob_bysex_byage.RDS")
```
