---
title: "GRAM Simulation"
author: "Sigal Maya"
date: "Updated 03/10/2025"
output:
  html_notebook
subtitle: Calibration and Benchmarking
knit_root_dir: /Users/smaya/Documents/GitHub/GRAM
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(knitr)
library(tableone)
library(kableExtra)
library(tibble)
library(patchwork)
```

```{r source files, include = FALSE}
# The following will only run properly when knit. While working on document, run this chunk in the console.
source("gram_model/gram_01setup.r")
source("gram_model/gram_02helpers.r")
source("gram_model/gram_03simulation.r")
source("gram_calibration/gram_benchmarking_helpers.r")
```

```{r benchmark data, include = FALSE}
benchmark_values <- tribble(
  ~Variable,                    ~`Benchmark Value`,     ~Source,
  "n",                          "",                     "",
  "AGE",                        "",                     "",
  "SEX: Male",                  "49.0%",                "ACS data",
  "SEX: Female",                "51.0%",                "ACS data",
  "RACEETH: Non-Hispanic White","64.0%",                "ACS data",
  "RACEETH: Non-Hispanic Black","14.0%",                "ACS data",
  "RACEETH: Hispanic",          "22.0%",                "ACS data",
  "EDU",                        "13.6 (??)",            "UNDP-HDI",
  "INCOME: Low",                "5.0%",                 "ACS data",
  "INCOME: Medium",             "17.0%",                "ACS data",
  "INCOME: High",               "78.0%",                "ACS data",
  "MEDBUR",                     "2.25 (??)",            "Mossadeghi et al 2023",
  "APOE4: Not Carrier",         "75.0%",                "Di Battista et al 2016",
  "APOE4: Carrier",             "25.0%",                "Di Battista et al 2016",
  "SYN: Cognitively Healthy",   "",                     ""
)

lifetable <- read.csv("gram_data/table_prob_die_next_year.csv") # these are probs, convert to rates
lifetable <- lifetable %>%
  mutate(rate = - log(1 - qx))

benchmark_prev_by_age <- data.frame(
  age = c(65, 70, 75, 80, 85, 90),
  raceeth = "Overall",
  condition = c(rep("benchmark_dem", times = 6), rep("benchmark_mci", times = 6)),
  prev = c(0.03, 0.04, 0.09, 0.18, 0.26, 0.35,
           0.22, 0.20, 0.21, 0.25, 0.22, 0.27)
)

benchmark_prev_by_race <- data.frame(
  age = rep(74, times = 3),
  raceeth = c("NHW","NHB","Hisp"),
  condition = c(rep("benchmark_dem", times = 3), rep("benchmark_mci", times = 3)),
  prev = c(0.11, 0.15, 0.10,
           0.23, 0.22, 0.28)
)

benchmark_reside_time <- data.frame(
  age_group = "Benchmark",
  condition = c("benchmark_mci", "benchmark_dem"),
  duration = c(3.85, 7)
)

# Note that the benchmark dementia duration is a guess based on conversations with Kelly and Kate.

benchmark_age_of_onset <- 71.5
```

#### **sim1: Initial model**

```{r sim1, include = TRUE}
l.inputs1 <- l.inputs # seed_stochastic is 20240202
l.inputs1[["n.cycle"]] <- 51  # in order to see age 100, since cycle = 1 is age 50
sample1 <- readRDS("gram_data/acs_data/acs_sample_1.rds")

sim1 <- run_benchmarking(l.inputs1, "sim1: original model", sample1)
```
Observations:
1. Modeled mortality is higher than benchmark, with increasing error with age. 
2. MCI prevalence is too low, dementia prevalence is too high.
3. Reside times in both MCI and dementia seem to be spot on.
4. Modeled mean age of onset is slightly older than benchmark (71.5 vs 72.3, which is about 1%).

Dementia prevalence is too high, which is likely causing the high mortality in older ages. The only way to get dementia in the model is to (a) get MCI, then (b) progression in CDR score. Increasing MCI incidence will push up dementia prevalence, reducing it will further reduce MCI prevalence (which is already lower than we want it to be). So I will first try to reduce CDR progression rate.

#### **sim2: Reduced CDR progression rate by 50% while in MCI**
```{r sim2}
l.inputs2 <- l.inputs1
l.inputs2[["r.CDRslow_mean"]] <- l.inputs1[["r.CDRslow_mean"]] * 0.5

sim2 <- run_benchmarking(l.inputs2, "sim2: reduce MCI CDR progression by 50%", sample1)
```
As expected, that did help lower mortality, increase MCI prevalence closer to benchmark, and decrease dementia prevalence closer to benchmark (though still a bit higher). But it also increased reside time. 
I need to lower mortality a little more, and also lower dementia prevalence. Try lowering MCI incidence (baseline / no risk factor) instead.

#### **sim3: Reduce baseline (reference) MCI hazard rate**

```{r}
l.inputs3 <- l.inputs1
l.inputs3[["m.hr_mci"]] <- l.inputs3[["m.hr_mci"]] * 0.5

sim3 <- run_benchmarking(l.inputs3, "sim3: lower MCI incidence", sample1)
```
Dementia prevalence seems acceptable but mortality at older ages is too low. Reduce dementia mortality adjustment for age - maybe we are already account for most of that in our estimation of non-dementia mortality.

#### **sim4: Lower MCI incidence (sim3) + no age adjustment for dementia-related mortality risk**
```{r}
l.inputs4 <- l.inputs3
l.inputs4[["hr.mort_age"]] <- c(1,0.8,0.6) # as opposed to 1, 0.6, 0.3 in initial model

sim4 <- run_benchmarking(l.inputs4, "sim4: lower MCI incidence + smaller dementia mortality adjustment for age")
```
Mortality seems acceptable now. MCI prevalence is super low, and mean age of onset is too old, because we reduced incidence rate. Model currently does not directly adjust for race/ethnicity. It looks like we underestimate dementia prevalence in Black people and overestimate among Hispanic people. A quick Google search tells me that African American people may be more likely to carry APOE4. Try adjusting the coefficients for those two groups to account for this.

#### **sim5: Introduce MCI hazard function coefficients for Black and Hispanic.**

```{r}
l.inputs5 <- l.inputs4
l.inputs5[["log_RACEETHblack"]] <- log(1.2)
l.inputs5[["log_RACEETHhisp"]] <- log(0.9)
  
sim5 <- run_benchmarking(l.inputs5, "sim5: sim4 + MCI hazard coefficients for race/ethnicity")
```
That fixed the differences across racial groups. But mean age of onset still to late, and MCI prevalence too low. Just as a test, I want to see what happens when I start with a prevalent population.

#### **sim6: sim5 + prevalent starting cohort**
```{r}
l.inputs6 <- l.inputs5
l.inputs6[["p.SYN_start"]] <- c(0.95, 0.05)
  
sim6 <- run_benchmarking(l.inputs6, "sim6: sim5 + prevalent cohort")
```
Not much changed, but it did bring down the average age of onset. It seems that the only way to have more MCI than dementia is to have people stay in MCI for longer. But, that means the model has longer reside times than benchmark. 

#### **sim7**
```{r}
l.inputs7 <- l.inputs6   # lower MCI incidence, lower age-impact of dementia mortality, MCI hazard coefs for race, prevalent cohort

l.inputs7[["r.CDRslow_mean"]] <- l.inputs7[["r.CDRslow_mean"]] * 0.5  # reduce CDR progression rate during MCI (increases reside time before converting to dementia)
# MCI prevalence too low, reside time too low

l.inputs7[["m.hr_mci"]] <- l.inputs1[["m.hr_mci"]]    # back to original MCI incidence
# MCI prevalence is better, but needs to be higher among Hisp

l.inputs7[["log_RACEETHhisp"]] <- log(1.3)            # original is 1.00
# MCI prevalence is good, dementia prevalence is too high, cumulative mortality too high especially at older age

l.inputs7[["hr.mort_age"]] <- l.inputs1[["hr.mort_age"]]
# dementia prevalence still too high, and mortality at older age still too high

l.inputs7[["m.lifetable"]] <- c(
  l.inputs7[["m.lifetable"]][1:21]  * 1.00,
  l.inputs7[["m.lifetable"]][22:51] * 0.95
)
# dementia prevalence is still too high, and cumulative mortality at older age is high

# l.inputs7[["r.CDRfast_mean"]] <- l.inputs7[["r.CDRfast_mean"]] * 0.75 
# that didn't do much other than reducing reside time in dementia, so will revert


sim7 <- run_benchmarking(l.inputs7, "sim7")
```

It's really hard to de-couple dementia prevalence from MCI prevalence. Wondering if a uniform random increase in CDR-SB is not a good strategy. For example, APOE4 carriers more likely to progress.
I think that's why it's not possible to get both MCI/dementia prevalence and reside times right. Once you get into MCI, you're in a set trajectory that doesn't allow for much variation (there is a standard deviation from the mean but that's fixed per individual and it's quite small). If we can have more granularity on the rate of progression to dementia (which will not be a random factor -- this is dependent on many things), I think we will be able to get closer to benchmarks on both areas.